---
title: "assignment"
author: "Gabriele Olendraite"
date: "2026-03-04"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub) # to fetch genomes, annotations, etc.
  library(ensembldb)     # to manipulate ensembl gene annotations
  library(GenomicRanges) # data structure for genomic ranges 
  library(ggplot2)
})
```

1. Using AnnotationHub, find and download the following annotations data:

    The mouse (Mus Musculus) EnsDb object, version 102, genome build GRCm38
    The mouse genome sequence ( dna_sm ) in TwoBit/2bit format for GRCm38
    The drosophila melanogaster genome sequence ( dna_sm ) in TwoBit/2bit format for BDGP6

```{r}
ah <- AnnotationHub()
```

```{r}
query(ah, c("Mus musculus", "EnsDb", "102", "GRCm38"))

```
```{r}
mm_ensdb <- ah[["AH89211"]]
```

```{r}
mcols(query(ah, c("Mus musculus", "dna_sm", "TwoBit", "GRCm38", "102")))
```
```{r}
mm_genome <- ah[["AH88477"]] 
```

```{r}
query_res <- query(ah, c("Drosophila melanogaster", "dna_sm", "TwoBit", "BDGP6"))
rev(query_res)[1]
```

```{r}
dm_genome <- ah[["AH106160"]]
```

2. Using the mouse EnsDb, find the following:
   How many different ensembl gene IDs and gene symbols are there for protein-coding genes?
   
```{r}
mm_genes = genes(mm_ensdb, filter=~gene_biotype=="protein_coding")
mm_genes
```

```{r}
length(unique(mm_genes$gene_id))
```
```{r}
length(unique(mm_genes$gene_name))
```

   
   
 Plot the distribution of the (spliced) length of protein-coding transcripts
 (tip: this will require you to extract exons of protein-coding transcripts from the database, and split them by transcript, before summing the width of the exons of each transcript)


```{r}
mm_exons <- exonsBy(mm_ensdb, by = "tx", filter=~tx_biotype=="protein_coding")
mm_exons
```

```{r}
exon_lengths <- width(mm_exons)

head(exon_lengths)
```
```{r}
total_tx_lengths <- sum(exon_lengths)

head(total_tx_lengths)
```

```{r}
df_sums <- data.frame(
  transcript_id = names(total_tx_lengths),
  total_length = as.numeric(total_tx_lengths)
)
```

```{r}
ggplot(df_sums, aes(x = total_length)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Spliced Protein-Coding Transcript Lengths",
       x = "Total Exon Length (bp)",
       y = "Number of Transcripts")
```
```{r}
ggplot(df_sums, aes(x = total_length)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  scale_x_log10() + 
  theme_minimal() +
  labs(title = "Distribution of Spliced Protein-Coding Transcript Lengths",
       x = "Total Exon Length (bp) - Log10 Scale",
       y = "Number of Transcripts")
```

